---
title: "R: Functions and Packages"
subtitle: "Part 4: Testing"
author: "Mik Black & Murray Cadzow"
date: "7 February 2018"
output: ioslides_presentation
---

## Attribution

 - These slides are modified for Murray Cadzow's 2017 "testing" session.
 - Much of this content is adapted from Hadley Wickham's R packages book:

    http://r-pkgs.had.co.nz/tests.html

 - There is also content from Karl Broman:

    http://kbroman.org/pkg_primer/pages/tests.html



## Getting set up

This will carry on with the example from lesson 1

```{r eval = FALSE}
install.packages('testthat')

library(testthat)
```

```{r echo=FALSE}
library(testthat)
```

change your working dir to 'cats'

```{r, eval = FALSE}
setwd("cats")
```

## Initialise the test directory

```{r, eval = FALSE}

devtools::use_testthat()

```

The `use_testthat()` command:

* creates a `tests/testthat` directory

* adds `testthat` to the "Suggests"" field in the DESCRIPTION file

* creates the file `tests/testthat.R` that runs all your tests are when `R CMD check` runs

## The "tests" directory

```{r, eval = FALSE}
list.files(path = "tests/", recursive=TRUE, include.dirs=TRUE)
```

* The `tests` directory contains `testthat.R` and `testthat/`

* We create our tests to be run in the `tests/testthat/` directory

* The filenames start with "test", e.g., `test_cat_function.R`

* The `tests/testthat.R` file tells `test` to: 
     - load the `testthat` library when testing
     - go through the tests that are stored in `tests/testthat/`

## Test structure

 - Test files contain:
    -  __context__: a description of what is being tested
 - One or more tests, organised hierarchically: 
    - __expectations__ are grouped into __tests__ which are organised in __files__

```{r, eval=FALSE}
library(stringr)
context("String length")

test_that("str_length is number of characters", {
  expect_equal(str_length("a"), 1)
  expect_equal(str_length("ab"), 2)
  expect_equal(str_length("abc"), 3)
})
```

## Tests

 - A test groups together multiple expectations to test output

 - This can range from a simple function through to a complex multi-argument function

 - Sometimes called a __unit__ test because it tests one unit of functionality


## Expectations

 - The 'atom' of testing

 - Describes the output of a computation

 - Does it produce what it should?

 - Does it produce an error code?

 - Expectations are functions that start with `expect_`

## Expectations continued 1

 - Expectations make a binary assertion about what a function does

 - Structure:
    - start with `expect_`
    - have two arguments, 1st: actual result, 2nd what was expected
    - if actual and expected don't match `testthat` produces an error

## Expectations continued 2

 - Some common expectations that we can use are:

```{r, eval = FALSE}
expect_equal()

expect_identical()

expect_match()

expect_output()

expect_message()

expect_error()
```

 - There are many others too

## Example: expect_equal

```{r, eval = TRUE, error=TRUE}
expect_equal(10, 10)

expect_equal(10, 10 + 1e-7)

expect_equal(10, 11)
```

 - Why didn't the second `expect_equal` call return an error?

## Example: expect_identical
 
 - `expect_equal` uses `all.equal()` which allows a numerical tolerance 

 - For exact comparisons use `expect_identical()`, which uses `identical()`

```{r eval = TRUE, error=TRUE}
expect_equal(10, 10 + 1e-7)

expect_identical(10, 10 + 1e-7)
```

## Example: strings

 - Note case sensitivity:
 
```{r, eval = TRUE, error=TRUE}
string <- "Testing is fun!"

expect_match(string, "Testing") 

expect_match(string, "testing")

expect_match(string, "testing", ignore.case = TRUE)
```

##  Example: matching

Variations of expect_match() allow you to check for:

 - Printed output: `expect_output()`
 - Messages: `expect_message()`
 - Warnings: `expect_warning()`
 - Errors: `expect_error()`

## What/when to test

 - If you find a bug, write a test __then__ fix

 - How can someone break your function?
    - What happens if an argument isn't passed?
    - What happens if someone passes the wrong data structure/type?
    - What happens with NAs?
    - Do your conditionals work as expected?
  
 - What situations does your function need to handle and what can you let fail?

## Cats example

 - Lets make a test for our cats function.
 
```{r, eval=FALSE}
devtools::use_testthat()
```

 - How would you check to make sure the function's output is correct?  
 - Write a function to test the output of `cats_function()`, and put it in a file called `testcats.R`
 in the `tests/testthat` directory.
 - Run your test via:
 
```{r, eval=FALSE}
devtools::test()
``` 
 
 